---
title: "NCAA Men's College Basketball Analysis"
author: "Jake Whalen"
date: "March 22, 2020"
output: 
    html_document:
      code_folding: hide
      number_sections: yes
      toc: yes
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.width = 8, fig.height = 3)
```

```{r packages, include=FALSE}
library(vroom)
library(tidyverse)
library(magrittr)
library(ggthemes)
library(gridExtra)
library(janitor)
library(ggrepel)
```

```{r helpers, include=FALSE}
my_theme <- 
    theme_solarized() +
    theme(panel.background = element_rect(color='black'))
```

# Purpose

The purpose of this project is to explore the Mens NCAA March Madness (MM) data for insight into what gives teams a winning advantage.

# Data

The data for this project was downloaded from the [Google Cloud & NCAA March Madness Analytics](https://www.kaggle.com/c/march-madness-analytics-2020/data) Kaggle competition.
The project will focus on the data from the NCAA Men's league.
The list of names below are the data files included for this analysis.
The data goes back to 1985.
The reason for cutting off at 1985 is because this was the first year the tournament expanded to 64 teams, similar to today but without the 4 play-in teams.

```{r data, cache=TRUE}
folder <- 'C:/Users/Jake/Documents/Projects/UVA-CBB/data/MDataFiles_Stage1'
files <- list.files(folder)
dfs <- list()

for (f in files){
    df.name <- str_replace(f, '.csv', '')
    dfs[[df.name]] <- vroom(file.path(folder, f))
}
files
```

In addition to the files above there are play-by-play records since 2015.
These files contain individual records for every event takig place during the season.
There is potential for the data to not be 100% accurate as described on the Kaggle site.

```{r pbp, cache=TRUE}
# folder <- 'C:/Users/Jake/Documents/Projects/UVA-CBB/data'
# files <- list.files(folder, pattern = "MEvents")
# pbp <- list()
# 
# for (f in files){
#     df.name <- str_replace(f, '.csv', '')
#     pbp[[df.name]] <- vroom(file.path(folder, f))
# }
# files
```


# Exploratory Data Analysis (EDA)

Some of the data files supplied contain useful metadata to go along with some of the other more detailed files.
These files include Cities, MGameCities, Conferences, MNCAATourneySeedRoundSlots, MSeasons, MTeamConferences, MNCAATourneySlots, MConferenceTourneyGames, and MTeamSpellings.
The data from these files will be incorporated into the analysis later on but there is not much to discover from these files alone.

## The Growth of a League

The __MTeams__ dataframe contains the TeamID, team name, and first/last season each team was active in.
There have been `r nrow(dfs$MTeams)` Division 1 (D1) men's teams that have played at least one seson since `r min(dfs$MTeams$FirstD1Season)` (first season recorded in data).
In `r min(dfs$MTeams$FirstD1Season)` there were `r nrow(filter(dfs$MTeams, FirstD1Season == min(dfs$MTeams$FirstD1Season)))` active D1 men's teams.
There were `r nrow(filter(dfs$MTeams, LastD1Season == 2020))` active teams in the most recent college basketball season (2019-2020).
The chart below shows the growth over time in the number of men's D1 basketball teams.
The chart shows a steady growth in the number of teams added up until the late 2000's.
After 2010 the growth has slowed down.

```{r}
left_join(dfs$MTeams %>%
              group_by(Season = FirstD1Season) %>% summarise(First = n()),
          dfs$MTeams %>%
              group_by(Season = LastD1Season) %>% summarise(Last = n()),
          by = 'Season') %>% 
    mutate(Last = ifelse(is.na(Last), 0, Last),
           fCS = cumsum(First),
           lCS = cumsum((Last)),
           current = fCS - lag(lCS),
           current = ifelse(is.na(current), First, current)) %>% 
    ggplot(aes(x=Season, y=current)) +
    geom_line() +
    geom_point(color = 'maroon') +
    labs(title = "Growth of NCAA Men's Div 1",
         y = 'Number of Teams', x = 'Season') +
    my_theme
```

## Regular Season



## March Madness

### Tournament Seeding

Selection Sunday is the Sunday on the final weekend of the regular season when conference tournaments wrap up.
The name comes from the fact that on that day the 64 (68 including play-ins) teams are selected on live TV to compete in MM.
While some teams are waiting to find out if they will be invited, a majority already know with almost near certainty that they will be given an invite.
Those teams who know they are in are mostly interested to see what their seed is and who their opponents will be.

Seeding in the tournament is very important.
It is widely considered that the higher the seed the better chance you have at making the Final Four and winning the National Championship.
The lowest seed to ever win the championship was Villanova at the 8 spot.
This isn't to say seeds below an 8 should be counted out, but history is not on their side.

#### Seed Matchup Performance

The graphic below shows the winnig percentage amongst seed matchups.
The seeds denoted on the y-axis are the winners while the seeds across the x-axis are the losers.
The diagonal (equal seeded matchups) were removed from this chart since there was no comparison (1 seed has a 100% win rate against 1 seeds with a 100% losing rate at the same time).
A blank indicates the two seeds have never matched up before in the tournament.
When viewing this chart it is important to remind yourself that not all of these matchups are as common as others.
For instance a 1 vs 16 and a 8 vs 9 are guaranteed to be played four times every single tournament because those are first round games.

A few quick takeaways from the graphic below:

* 1 Seeds:
    + Have lost to a 16 only one time in the history of the tournament.
    + As expected, have a greater than 50% winning percentage against all other seeds.
    + Have never faced a 14 or 15 seed due to these lower seeds never advancing far enough (Elite 8).
    + Struggle with the 11 seeds, only posting a 57% winning percentage.
* Being a 2 seed doesn't look as good as it sounds.
    + 0 wins over 5 and 9 seeds.
    + An abysmal 29% winning percentage over 8 seeds.
* In recent memory the 5 vs 12 is a common upset pick in many brackets however the 12 seed only holds a 36% winning percentage in this matchup.


```{r, fig.height=6}
df.seed.results <-
    dfs$MNCAATourneyCompactResults %>% 
    select(Season, DayNum, WTeamID, LTeamID) %>% 
    left_join(., 
              dfs$MNCAATourneySeeds %>% rename(WTeamID = TeamID, WSeed = Seed),
              by = c('Season', 'WTeamID')) %>% 
    left_join(., 
              dfs$MNCAATourneySeeds %>% rename(LTeamID = TeamID, LSeed = Seed),
              by = c('Season', 'LTeamID')) %>% 
    mutate(WSeed = str_replace(WSeed, 'a', ''),
           WSeed = str_replace(WSeed, 'b', ''),
           LSeed = str_replace(LSeed, 'a', ''),
           LSeed = str_replace(LSeed, 'b', ''),
           WReg = substr(WSeed, 1, 1),
           LReg = substr(LSeed, 1, 1),
           W_L_Regions = paste(WReg, LReg, sep = "/"),
           WSeed = as.numeric(substr(WSeed, 2, 3)),
           LSeed = as.numeric(substr(LSeed, 2, 3)))

df.wvl.seed <-
    df.seed.results %>%
    group_by(WSeed, LSeed) %>% 
    summarise(WSeedWins = n())

df.wvl.seed %>% 
    left_join(., 
              df.wvl.seed %>% 
                  rename(LSeed = WSeed, WSeed = LSeed,
                         LSeedWins = WSeedWins),
              by = c('WSeed', 'LSeed')) %>%
    filter(WSeed != LSeed) %>% 
    mutate(LSeedWins = ifelse(is.na(LSeedWins), 0, LSeedWins),
           WSeedWinPct = WSeedWins / (WSeedWins + LSeedWins)) %>%
    ggplot(aes(x=LSeed, y=WSeed, fill=WSeedWinPct)) +
    geom_tile() +
    geom_text(aes(label = round(WSeedWinPct, 2))) +
    geom_segment(aes(x=0.5, y=0.5, xend=16.5, yend=16.5)) +
    geom_hline(yintercept = 0.5:16.5) +
    geom_vline(xintercept = 0.5:16.5) +
    scale_y_continuous(trans = "reverse", breaks = 1:16, expand = c(0,0)) +
    scale_x_continuous(position = "top", breaks = 1:16, expand = c(0,0)) +
    scale_fill_gradient2(low = 'red', mid='grey',
                         high='green', midpoint = 0.5) +
    my_theme +
    theme(axis.ticks = element_blank(),
          legend.position = 'bottom',
          panel.grid.major = element_blank()) +
    labs(x = 'Losing Team Seed', y = 'Winning Team Seed')
```

#### Average Seed vs MM Winning Percentage

The graph below plots each teams average tournamnet seed (x-axis) against their tournamnet winning percentage all-time (y-axis).
The graph illustrates what most would expect, higher seeds (lower on x-axis) win more games in the tournament (higher on the y-axis).
You see the blue bloods in the top left corner as expected.
Loyola-Chicago seems to be a bit of an outlier.
This is primarily due to their Final Four run in 2018 as an 11 seed.
Prior to that they had been to the tournament one time as a 4 seed in 1985.
On the other end there is a team like TCU who has consistently underperformed there seeding.
With three trips to the tournament as a 4, 5, and 6 seed they have posted a sub 30% winning percentage.

```{r, fig.height=6}
team.seeds <-
    dfs$MNCAATourneySeeds %>% 
    mutate(Seed = str_replace(Seed, 'a', ''),
           Seed = str_replace(Seed, 'b', ''),
           Region = substr(Seed, 1, 1),
           Seed = as.numeric(substr(Seed, 2, 3))) %>% 
    left_join(., 
              dfs$MTeams %>% select(TeamID, TeamName))

team.seeds.summ <-
    team.seeds %>% 
    group_by(TeamName) %>% 
    summarise(Appearances = n(), AvgSeed = mean(Seed),
              High = max(Seed), Low = min(Seed)) %>% 
    ungroup()

#Avg Seed vs Winning %
#filter(dfs$MNCAATourneySeeds, TeamID == filter(dfs$MTeams, TeamName == 'Cleveland St')$TeamID)
dfs$MNCAATourneyCompactResults %>%
    group_by(TeamID = WTeamID) %>%
    summarise(Wins = n()) %>% 
    left_join(.,
              dfs$MNCAATourneyCompactResults %>%
                  group_by(TeamID = LTeamID) %>%
                  summarise(Losses = n()),
              by = 'TeamID') %>% 
    left_join(., 
              dfs$MTeams %>% select(TeamID, TeamName),
              by = 'TeamID') %>% 
    mutate(WinPct = Wins / (Wins + Losses)) %>% 
    select(TeamName, WinPct) %>% 
    left_join(., team.seeds.summ, by = 'TeamName') %>% 
    ggplot(aes(x=AvgSeed, y=WinPct)) +
    geom_text(aes(label = TeamName, color = TeamName),
              size = 3, show.legend = F) +
    my_theme +
    labs(y = 'MM Winning Percentage', x = 'Average Seed') +
    scale_y_continuous(labels = scales::percent)
```

### Tournament History of Individual Teams

The graphs below take a look at the frequency with which teams made the tournament over the past 35 seasons (left) and their average seed when making the tournament (right).
These graphics confirm what most who follow college basketball already knew, the blue bloods (Duke, UNC, UK, KU) are nicknamed that for a reason.
The consistently make the tournament and they are consistently given high seeds.


```{r, fig.height=4}
#Teams history of being seeded
p1 <- team.seeds.summ %>% 
    ungroup() %>% 
    mutate(rank = rank(-Appearances, ties.method = 'min')) %>% 
    filter(rank <= 10) %>% 
    ggplot(aes(x = Appearances, y=reorder(TeamName, Appearances))) +
    geom_bar(stat = 'identity', fill = 'lightblue', color = 'black') +
    geom_text(aes(x=Appearances/2, label = Appearances)) +
    my_theme +
    labs(y = 'Team', x = 'Tournament Appearances (Max 35)')

#Teams average seed
p2 <- team.seeds.summ %>% 
    mutate(rank = rank(AvgSeed, ties.method = 'min')) %>% 
    filter(rank <= 10) %>% 
    ggplot(aes(x = AvgSeed, y=reorder(TeamName, -AvgSeed))) +
    geom_bar(stat = 'identity', fill = 'lightblue', color = 'black') +
    geom_text(aes(x=AvgSeed/2, label = sprintf("%s (%s)", round(AvgSeed, 2), Appearances))) +
    my_theme +
    labs(y = 'Team', x = 'Average Seed (Appearances)')

grid.arrange(p1, p2, ncol=2)
```

### Team Conferences 

#### Conference Matchup Performance MM

Since 1985 there have been a total of 51 conferences in the college basketball landscape.
The chart below shows how each conference has faired in tournament games against other conferences.
In order to make the graphic below more legible, only the results for conferences still competing in 2019-20 are displayed.
Each row contains the conferences winning percentage against all other conferences.
For teams that moved conferences (i.e. Syracuse or Louisville) their results are applied to the conference that they belonged to at the time.

A few observations on the results of the chart below:

* The ACC, Big Ten, Big Twelve, and SEC post very few losing records against all other conferences. Combined they hold losing records against only 7 other conferences.
* The ACC has a really poor reputation against the Horizon league suprisingly.
* CUSA has a perfect record against 13 other conferences. This is most likely a factor of a smaller sample size rather than actual dominance in the tournament.
* The SWAC is better to bet against, with 14 zero win records against other conferences. This again is less likely to be an accurate depiction of the conferences true performance and more a case of very few games in the tournament.

```{r, fig.height=8}
#Conf vs conf
cnf_v_cnf <-
    dfs$MNCAATourneyCompactResults %>% 
    select(Season, WTeamID, LTeamID) %>% 
    left_join(.,
              dfs$MTeamConferences %>%
                  rename(WTeamID = TeamID, WConf = ConfAbbrev)) %>% 
    left_join(.,
              dfs$MTeamConferences %>%
                  rename(LTeamID = TeamID, LConf = ConfAbbrev)) %>% 
    group_by(WConf, LConf) %>% 
    count()

keep_conf <-
    dfs$MTeamConferences %>% 
    filter(Season == 2020) %>%
    group_by(ConfAbbrev) %>% 
    count() %>% 
    select(Conf = ConfAbbrev) %>% 
    .[[1]]

cnf_v_cnf %<>%
    filter(WConf %in% keep_conf, LConf %in% keep_conf)

cnf_v_cnf %>% 
    full_join(.,
              cnf_v_cnf %>% 
                  rename(WConf = LConf,
                         LConf = WConf,
                         n_L = n)) %>%
    filter(WConf != LConf) %>% 
    mutate(n = ifelse(is.na(n), 0, n),
           n_L = ifelse(is.na(n_L), 0, n_L),
           WinPct = n / (n + n_L)) %>% 
    ggplot(aes(y=WConf, x=LConf, fill=WinPct)) +
    geom_tile() +
    geom_text(aes(label = round(WinPct, 2)), size=2, angle = 35) +
    geom_segment(aes(x=0.5, y=32.5, xend=32.5, yend=0.5)) +
    geom_hline(yintercept = 1.5:32.5) +
    geom_vline(xintercept = 1.5:32.5) +
    scale_x_discrete(position = 'top', expand = c(0,0)) +
    scale_y_discrete(limits = rev(unique(cnf_v_cnf$WConf)), expand = c(0,0)) +
    scale_fill_gradient2(low = 'red', mid='grey',
                         high='green', midpoint = 0.5) +
    my_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = -.1),
          panel.grid.major = element_blank(),
          legend.position = 'bottom') +
    labs(x='Losing Team Conference', y='Winning Team Conference')
```

#### Conference Success

##### Number of Bids

##### Number of Championships

## Coaches

MTeamCoaches

## Regular Season Rankings

MMasseyOrdinals

# Asking the Data Questions

* Coaching changes
* Similarity between reg season and playoffs
* Distance from home in tourney
* Best region to be in to win it all?
* Do regular season matchups determine MM outcome?


# Conclusions

## What gives Teams an Advantage?